Anxiety: a decision-theoretic perspective

### 2019-02-02 manuscript/manuscript.tex
minor discussion updates

An emerging idea in this literature is that pathological avoidance as a result
of anxiety mediates the transition to depression, perhaps by reduce the likelihood
of experiencing positive events and activities \citep{moitra2008, jacobson2014}.
This finding is perfectly in line with the model presented so far. A sketch of
how this is possible is shown in Figure 6. Under unperturbed learning, avoidable
threat does not impede behavior. Under pessimistic learning, threat can block
future reward. This avoidance can manifest as a failure of behavioral initiation,
or a lack of overall mobility (here represented by an agent self-transitioning
at the starting state). Thus, the model can accomodate the deficits in positive
affect observed in depression as a function of behavioral avoidance, as predicted
by these epidemiological studies.

@@

In this article, we showed...

Discuss previous computational approaches: This has also been previously been
investigated in previous computational modeling \citep{HuysDayan2009}. There
the authors found that beliefs about a lack of controllability, mechanistically
modeled as corrupted priors about action-outcome contingencies, found deficient
learning in novel environments. The present results are no doubt related, but
address a different environment and a different behavior.

@@

\section{References}

>>>>

An emerging idea in this literature is that pathological avoidance as a result
of anxiety mediates the transition to depression, perhaps by reduce the likelihood
of experiencing positive events and activities \citep{moitra2008, jacobson2014}.
This finding is perfectly in line with the model presented so far. A sketch of
how this is possible is shown in Figure 6. Under unperturbed learning, avoidable
threat does not impede behavior. Under pessimistic learning, threat can block
future reward. This avoidance can manifest as a failure of behavioral initiation,
or a lack of overall mobility (here represented by an agent self-transitioning
at the starting state). Thus, the model can accommodate the deficits in positive
affect observed in depression as a function of behavioral avoidance, as predicted
by these epidemiological studies.

@@

All of the anxiety disorders share core symptoms including pessimistic inference,
second-order conditioning, and sustained avoidance. In this article, we offered
a simple computational account of how these aberrations may arise and be maintained.
Following the literature in Markov decision processes, we showed how a simple
breakdown in expectations about the likelihood of encountering future negative
outcomes (as a result of an unpredictable environment or unreliable policy) can
effectively backpropagate negative expectations through an agent's internal value
map. The result of which yields expectations and behaviors resembling that observed
in pathological anxiety. Importantly, we also showed that such a model can account
for aberrant behavior previously observed in individuals with anxiety in several
assays of psychiatric behavior. By no means a complete theory of anxiety, this
account helps to at least tie together several core symptoms of anxiety in one
account.

This computational account draws upon a longstanding recognition of the importance
of perceived control to the development and maintenance of anxiety disorders. Several
prominent theories of anxiety in the clinical anxiety focus on the importance of
a lack of perceived control, including helplessness-hopelessness theories, perceived
efficacy, and the triple vulnerability model. It is worth noting that that these
theories differ, among other things, in where a lack of control arises. For learned
helplessness, an agent believes to occupy an uncontrollable world; in other words,
there is no reliable mapping between action and outcome. For self-efficacy theory,
an agent believes to be limited or incapable of exerting beneficial influence on
the environment, though this is otherwise possible. Note that both can easily be
modeled in the current approach, either in the state transition function or in
future policy function. Insofar that they are multiplicative, they should result
in the same effect. Future work may tease these apart, as they may be interesting
for guiding treatment.

It is worth noting that we are not the first to attempt to formalize a theory
of control in MDP. \cite{HuysDayan2009} provided a first computational account
of learned helplessness through simple models of action-outcome contingencies.
\cite{HuysDayan2009} found that training an agent first in an unreliable environment
led to later impaired learning, akin to early learned helplessness experiments,
by virtue of the prior of controllability. This provides a nice demonstration
of learning impairment of action-outcome contingencies. Importantly, our accounts
in at least three respects. We handle sequential decision environments, which is
important as it is in sequential decision environments where the consequences
of pessimistic inference become most pronounced. Second, our account claims
that the worst outcomes are believed, wherein their account focuses on randomness.
This seems closer to the symptoms described in anxiety. Finally, our account
provides insights into maintenance. With an otherwise unmodified Bayesian learner,
their agents should eventually learn and be ok.

@@
